{\Large \bf Analysing time course microarray data using Bioconductor: a case study using yeast2 Affymetrix arrays}

### 2016-01-25 paper.Rnw

<<echo=FALSE>>=
opts_chunk$set(prompt = FALSE,fig.path='graphics/', fig.align='center',fig.lp="")
#knit_hooks$set()
options(width=75)
knit_hooks$set(
    par.nice = function(before, options, envir) {
        if (before && !is.null(options$par.mfrow)){
            par(mar=c(3,3,2,1),  mgp=c(2,0.4,0), tck=-.01,
                cex.lab=.95,cex.axis=0.9, las=1, mfrow=options$par.mfrow)

@@

<<eval=FALSE,tidy=FALSE>>=
##From Bioconductor
biocLite(c('ArrayExpress', 'Mfuzz', 'timecourse', 'yeast2.db', 
    'yeast2probe', 'yeast2cdf'))
##From cran
install.packages(c('GeneNet', 'gplots'))
@

@@

<<eval=TRUE,cache=TRUE,tidy=FALSE>>=
ph = yeast.raw@phenoData
exp_fac = data.frame(data_order = 1:30, 
    strain = ph@data$Factor.Value..GENOTYPE.,  
    replicates = ph@data$Factor.Value..INDIVIDUAL.,
    tps = ph@data$Factor.Value..TIME.)
levels(exp_fac$strain) = c('m', 'w')
exp_fac = with(exp_fac, exp_fac[order(strain, replicates, tps), ])
exp_fac$replicate = rep(1:3, each=5, 2)
@

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=TRUE>>=
for(i in 1:5) {
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<s_fig2,eval=FALSE,tidy=FALSE>>=
d = exp_fac$data_order[1:5]
hist(yeast.raw[,d], lwd=2, ylab="Density", xlab="Log (base 2) intensities") 
@
\noindent produces the image shown in Appendix B: Figure \ref{F2A}. Typically,
differences in spread and position are corrected by normalisation. However, the
appearance of significant multi-modality in the distribution or many outlying
observations are indicative of poor data quality.

@@

<<s_fig3,eval=FALSE>>=
library(affyPLM)
#Raw data intensities
boxplot(yeast.raw, col='red', main='', ylim=c(2,16))
#Normalised intensities
boxplot(yeast.rma, col='blue', ylim=c(2,16))
@

@@

This package assesses treatment differences by comparing time-course
mean profiles allowing for variability both within and between time
points. It uses the multivariate empirical Bayes model proposed by
\cite{Tai06}. Further details of the \texttt{timecourse} package can be
found in \cite{Tai07}. After installing the \texttt{timecourse} package,
we construct a \texttt{size} matrix describing the replication
structure using
<<eval=TRUE,message=FALSE>>=
library(timecourse)
size = matrix(3, nrow = 5900, ncol = 2)
@
\noindent To extract a list of differentially expressed we calculate the
Hotelling statistic $\tilde T^2$ via
<<eval=TRUE,tidy=FALSE, cache=TRUE>>=
c.grp = as.character(exp_fac$strain)
t.grp = as.numeric(exp_fac$tps)
r.grp = as.character(exp_fac$replicate)
MB.2D = mb.long(yeast.matrix, times = 5, method = '2', 
    reps = size, condition.grp = c.grp, time.grp = t.grp, 
    rep.grp = r.grp)
@
\noindent The top (say) one hundred genes can be extracted via
<<eval=TRUE,cache=TRUE>>=
gene_positions = MB.2D$pos.HotellingT2[1:100]
gnames = rownames(yeast.matrix)
gene_probes = gnames[gene_positions]
@
\noindent The expression profiles can also be easily obtained. The profile for the top ranked expression is found using
<<s_fig4,eval=FALSE>>=
plotProfile(MB.2D, ranking=1, gnames=rownames(yeast.matrix))
@

@@

<<eval=TRUE,tidy=FALSE>>=
library(limma)
expt_structure = factor(colnames(yeast.matrix))

@@

<<F1A,eval=TRUE,dev="png",fig.ext="png",cache=TRUE,dpi=300,out.width="0.8\\linewidth",echo=FALSE, fig.cap="Image plots of the mismatch and perfect match probe intensities for the first replication of the mutant yeast strain. The corresponding times are indicated in the plot.", fig.pos="h">>=
par(mar=c(3,3,2,1), 
           mgp=c(2,0.4,0), tck=-.01,
               cex.axis=0.9, las=1, mfrow=c(2,3))
run_chunk("s_fig1")
@

@@

<<F2A,eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.6\\linewidth",echo=FALSE, fig.cap="Density plots for the first replication of the mutant yeast strain.",nice.par=TRUE,message=FALSE>>=
par(mar=c(3,3,2,1), 
           mgp=c(2,0.4,0), tck=-.01,
               cex.axis=0.9, las=1)
run_chunk("s_fig2")
@

<<F3A,eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.8\\linewidth",echo=FALSE, message=FALSE,fig.cap="Boxplots of the raw and normalised intensities. The default boxplot is to include both PM and MM intensities, whereas for the density plots in Figure~\ref{F2} the default is for only the PM intensities.">>=
par(mar=c(3,3,2,1), 
           mgp=c(2,0.4,0), tck=-.01,
               cex.axis=0.9, las=1, mfrow=c(1,2))
run_chunk("s_fig3")
@

\bigskip
<<F4A,eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.6\\linewidth",echo=FALSE, fig.cap="Time course expression levels for the top differentially expressed gene, ranked by their Hotelling statistic using the \texttt{timecourse} library.">>=
par(mar=c(3,3,2,1), 
           mgp=c(2,0.4,0), tck=-.01,
               cex.axis=0.9, las=1)
run_chunk("s_fig4")
@

>>>>

<<echo=FALSE>>=
library("knitr")
opts_chunk$set(prompt = FALSE,fig.path='graphics/', fig.align='center',fig.lp="")
#knit_hooks$set()
options(width=75)
knit_hooks$set(
    par.nice = function(before, options, envir) {
        if (before && !is.null(options$par.mfrow)){
            par(mar=c(3,3,2,1),  mgp=c(2,0.4,0), tck=-.01,
                cex.lab=.95,cex.axis=0.9, las=1, mfrow=options$par.mfrow)

@@

<<eval=FALSE,tidy=FALSE>>=
##From Bioconductor
biocLite(c('ArrayExpress', 'Mfuzz, 'timecourse', 'yeast2.db', 
    'yeast2probe', 'yeast2cdf', aAffyPLM'))
##From cran
install.packages(c('GeneNet', 'gplots'))
@

@@

<<eval=TRUE,cache=TRUE,tidy=FALSE>>=
ph = yeast.raw@phenoData
exp_fac = data.frame(data_order = 1:30, 
    strain = ph@data$Factor.Value..phenotype.,  
    replicates = ph@data$Characteristics..individual.,
    tps = ph@data$Factor.Value..time.)
levels(exp_fac$strain) = c('m', 'w')
exp_fac = with(exp_fac, exp_fac[order(strain, replicates, tps), ])
exp_fac$replicate = rep(1:3, each=5, 2)
@

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=TRUE, echo=-1>>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1, mfrow=c(2,3))
for(i in 1:5) {
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<s_fig2,eval=FALSE,tidy=FALSE,echo=-1>>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1)
d = exp_fac$data_order[1:5]
hist(yeast.raw[,d], lwd=2, ylab="Density", xlab="Log (base 2) intensities") 
@
\noindent produces the image shown in Appendix B: Figure \ref{F2A}. Typically,
differences in spread and position are corrected by normalisation. However, the
appearance of significant multi-modality in the distribution or many outlying
observations are indicative of poor data quality.

@@

<<s_fig3,eval=FALSE, echo=-1 >>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01,cex.axis=0.9, las=1, mfrow=c(1,2))
library(affyPLM)
#Raw data intensities
boxplot(yeast.raw, col='red', main='', ylim=c(2,16))
#Normalised intensities
boxplot(yeast.rma, col='blue', ylim=c(2,16))
@

@@

This package assesses treatment differences by comparing time-course
mean profiles allowing for variability both within and between time
points. It uses the multivariate empirical Bayes model proposed by
\cite{Tai06}. Further details of the \texttt{timecourse} package can be
found in \cite{Tai07}. After installing the \texttt{timecourse} package,
we construct a \texttt{size} matrix describing the replication
structure using
<<eval=TRUE,message=FALSE>>=
library(timecourse)
size = matrix(3, nrow = 5900, ncol = 2)
@
\noindent To extract a list of differentially expressed we calculate the
Hotelling statistic $\tilde T^2$ via
<<eval=TRUE,tidy=FALSE, cache=TRUE>>=
c.grp = as.character(exp_fac$strain)
t.grp = as.numeric(exp_fac$tps)
r.grp = as.character(exp_fac$replicate)
MB.2D = mb.long(yeast.matrix, times = 5, method = '2', 
    reps = size, condition.grp = c.grp, time.grp = t.grp, 
    rep.grp = r.grp)
@
\noindent The top (say) one hundred genes can be extracted via
<<eval=TRUE,cache=TRUE>>=
gene_positions = MB.2D$pos.HotellingT2[1:100]
gnames = rownames(yeast.matrix)
gene_probes = gnames[gene_positions]
@
\noindent The expression profiles can also be easily obtained. The profile for the top ranked expression is found using
<<s_fig4,eval=FALSE, echo=-1>>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1)
plotProfile(MB.2D, ranking=1, gnames=rownames(yeast.matrix))
@

@@

<<eval=TRUE,tidy=FALSE, message=FALSE>>=
library(limma)
expt_structure = factor(colnames(yeast.matrix))

@@

<<s_fig11,ref.label="s_fig1",eval=TRUE,dev="png",fig.ext="png",cache=FALSE,dpi=300,out.width="0.8\\linewidth",echo=FALSE, fig.cap="Image plots of the mismatch and perfect match probe intensities for the first replication of the mutant yeast strain. The corresponding times are indicated in the plot.", fig.pos="h">>=
#knitr::run_chunk("s_fig1")
@

@@

<<F2A,ref.label="s_fig2",eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.6\\linewidth",echo=FALSE, fig.cap="Density plots for the first replication of the mutant yeast strain.",nice.par=TRUE,message=FALSE>>=
@

<<F3A,ref.label="s_fig3",eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.8\\linewidth",echo=FALSE, message=FALSE,fig.cap="Boxplots of the raw and normalised intensities. The default boxplot is to include both PM and MM intensities, whereas for the density plots in Figure~\ref{F2} the default is for only the PM intensities.">>=
@

\bigskip
<<F4A,ref.label="s_fig4", eval=TRUE,dev="pdf",fig.ext="pdf",cache=TRUE,out.width="0.6\\linewidth",echo=FALSE, fig.cap="Time course expression levels for the top differentially expressed gene, ranked by their Hotelling statistic using the \texttt{timecourse} library.">>=
@
